---
title: 'NAFEW 2026 R Workshop: GRTS with TEON/AIM'
output: html_document
date: "2026-04-28"
---

```{r setup}
library(tidyverse) # general use
library(readxl)    # read in excel files
library(sf)        # spatial objects and analysis
library(spsurvey)  # spatial sampling design (GRTS)
# library(leaflet)   # interactive mapping/visualization
```

# TEON Use Case

## Load Data 

```{r}
# set filepath for all data
path = "C:/Users/ShaleHunter/OneDrive - USDA/Documents/TEON-misc/Conferences/NAFEW2026/GRTS_Workshop_Files"
```

```{r TEONdata}

# pre-cleaned historical data selection
TEON_info = read_csv(paste0(path, "TEON_project_info_NAFEW2026.csv"))

# examine data to see what we're working with
view(TEON_info)
```

## Data Preformatting

```{r TEONformatting}
# filter to useful subset
TEON_input_sites = TEON_info |> 
      # filter to sites with historical PCs - excludes some LTUB sites  
      # V10 is missing historical PC but needs to be included because it was already sampled in our first year
  filter(!is.na(hist_pc_data) | site == "V10") |>
  # grts() needs non-legacy sites to be NA
  mutate(selected_in_30_pairs = if_else(selected_in_30_pairs == "YES",
                                        "YES",
                                        NA))

# take a look at changes
view(TEON_input_sites)

# make sure all our already sampled sites are included
summarise(TEON_input_sites,
          .by = selected_in_30_pairs,
          n = n())

# note elevation disparity - will be important later
summarise(TEON_input_sites,
          .by = elevation_7500ft,
          n = n())

# format spatial
sf_input_sites = st_as_sf(TEON_input_sites,
                          coords = c("utm_e_z10", "utm_n_z10"),
                          # NAD27
                          crs = 26710,
                          remove = FALSE)

# visual check
plot(sf_input_sites)
```

## Running GRTS

```{r GRTSinfo}
# inspect GRTS function
?grts()
```

#### Documentation notes:

*stratum* vs *caty*:

stratification is for equal inclusion probability across groups
      ex: want to say something about each habitat type in a watershed, so same number of sites in forest/shrub/grass
      
category selection allows unequal sampling based on auxiliary variable
      ex: much more forest in watershed than shrub/grass, so want more forest sites
      
*legacy_var*:

the variable that describes which sites are "legacy" (already part of an existing sample design/already have data collected)

*n_over* and *n_near*:

In case some of your base sites must be rejected (terrain, access, etc) how do you want to fill in the gaps they leave in your design?

*n_over* selects sites in reverse hierarchical order
*n_near* uses nearest neighbor replacement sites

```{r GRTSdesign}
### generate GRTS sample

# ensure even elevation split
caty_n = c(H = 50, L = 50)

# set seed for reproducibility 
set.seed(330) 

# GRTS function
TEON_GRTS = grts(sframe = sf_input_sites,
                   n_base = 100,
                   legacy_var = "selected_in_30_pairs",
                   caty_n = caty_n,
                   caty_var = "elevation_7500ft")

plot(TEON_GRTS$sites_base)

```


```{r GRTSoutput}
# examine pieces of GRTS output
# legacy sites
view(TEON_GRTS$sites_legacy)
# base (new) sites
view(TEON_GRTS$sites_base)
```

#### GRTS output highlights:

*siteID* provides the **order** of sites - (remember this is important for maintaining the statistical validity of the design if sites are missed)

*wgt* and *ip*:
design weight and inclusion probability (ip is inverse of wgt). ip is high for high elevation sites (94% - we wanted 50 H sites, and only 53 were available from our historical site pool) and lower for low elevation sites (37% - we wanted 50 L sites, 135 available).



```{r GRTSinspection}
### check spatial balance using sp_balance()

# already sampled sites
sp_balance(object = TEON_GRTS$sites_legacy, 
           sframe = sf_input_sites)
# new sampling sites
sp_balance(object = TEON_GRTS$sites_base, 
           sframe = sf_input_sites)

# combine base and legacy 
fulldesign_TEON = rbind(TEON_GRTS$sites_legacy, TEON_GRTS$sites_base)

sp_balance(object = fulldesign_TEON, 
           sframe = sf_input_sites)


### summarize by other variables of interest to inspect how well GRTS sampled the available sites by non-spatial metrics

# use sp_summary() to quickly summarize GRTS output by variable
sp_summary(TEON_GRTS, ~ elevation_7500ft)

# can inspect individual variables independently
sp_summary(TEON_GRTS, ~ elevation_7500ft + dev_gradient + side + project)

# as well as the intersection of multiple variables
sp_summary(TEON_GRTS, ~ elevation_7500ft * side)

# need to refer back to distribution of full site list to compare representation
summarize(TEON_input_sites,
          .by = side,
          n = n())

summarize(TEON_input_sites,
          .by = c(side, elevation_7500ft),
          n = n()) |>
  arrange(side) # arrange by side to more easily compare to output of sp_summary()
  
```


```{r TEONDataExport}
#| eval = FALSE

# Export to csv and shp
write_csv(fulldesignTEON, paste0(path, "/NAFEW2026_GRTSdesign_TEON.csv"))
st_write(fulldesignTEON, paste0(path, "/NAFEW2026_GRTSdesign_TEON.shp"), append = FALSE)
```


# AIM Use Case

## Load Data

```{r AIMdata}

# shapefile of study area (BLM field office)
BFO = st_read(paste0(path, "/BFO_BLM_SMA/BFO_BLM_SMA.shp"))

# take a look at the AOI
plot(BFO)

```

## Data Preformatting

```{r}
# Nothing to see here!
```

## Running GRTS

```{r GRTSdesignAIM}
### generate GRTS sample

# set seed for reproducibility 
set.seed(330) 

# GRTS function
AIM_GRTS = grts(sframe = BFO,
                   n_base = 100,
                   n_over = 40
                 )

plot(AIM_GRTS$sites_base)

```

```{r GRTSoutputAIM}
# examine pieces of GRTS output
view(AIM_GRTS)

# base sites
view(AIM_GRTS$sites_base)

# oversample sites
view(AIM_GRTS$sites_over)
```

#### GRTS output highlights:

*sites_legacy* still exists in the structure of the GRTS output - its just empty because we aren't using any legacy sites for this sample frame.

*siteuse* distinguishes between base ("Base") and oversample ("Over") sites.

*ip*: inclusion probability is much smaller when you have an entire polygon to pull from instead of a set number of historical site locations.


```{r GRTSinspectionAIM}
### check spatial balance using sp_balance()

# check balance for base sites
sp_balance(object = AIM_GRTS$sites_base, 
           sframe = BFO)

# combine base and oversample sites
fulldesign_AIM = rbind(AIM_GRTS$sites_base, AIM_GRTS$sites_over)

# check balance for base and oversample sites together
sp_balance(object = fulldesign_AIM, 
           sframe = BFO)

```

```{r SpatialCheck}
# Because points can be selected from anywhere within BFO, there is more room for sites that may need to be rejected (for example, because of overlap with non-BLM managed lands):

# create buffer
BFO_edge_buffer = st_boundary(BFO) |>
  st_buffer(dist = 50) 

# find sites too close to allotment boundaries
grts_edge_plots = st_intersection(fulldesign_AIM,
                                  BFO_edge_buffer)

# remove marginal sites from GRTS sample frame
verified_design_AIM = filter(fulldesign_AIM,
                             !siteID %in% grts_edge_plots$siteID)

view(verified_design_AIM)

```

```{r AIMDataExport}
#| eval = FALSE

# Export to csv and shp
write_csv(verified_design_AIM, paste0(path, "/NAFEW2026_GRTSdesign_AIM.csv"))
st_write(verified_design_AIM, paste0(path, "/NAFEW2026_GRTSdesign_AIM.shp"), append = FALSE)
```


### Wrap-Up Notes

Vignettes from the **spsurvey** package: https://usepa.github.io/spsurvey/articles/start-here.html


